---
title: "NanoImprint_v2.0"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  html_document:
#params:
#  SAMPLE: ""
#  UNPHASED: ""
#  H1: ""
#  H2: ""
#  CONTROL: ""
#  REF: ""
#  REGIONS: ""
---

#### Sample: **<span style="color:blue">`r snakemake@params[["sample"]]`</span>** \

```{r include=FALSE}
library(xlsx, warning(F))
library(tidyverse, warning(F))
require(gridExtra)
library(ggplot2, warning(F))
library(edgeR, warning(F))
library(knitr)
library(kableExtra)
```



```{r Import sample data, echo=FALSE}
#Command line
sample_df <- read.delim(snakemake@input[["full"]], header=F) %>%
  filter(V4 == "m")
hp1 <- read.delim(snakemake@input[["H1"]], header=F) %>%
  filter(V4 == "m")
hp2 <- read.delim(snakemake@input[["H2"]], header=F) %>%
  filter(V4 == "m")
```



```{r Import control data, include=FALSE}
#Command line
ctrls= read.xlsx(snakemake@input[["control"]],1)
ctrls<- ctrls [, c(2:5)]

#dataset with selected cloumns
sorted_df <- sample_df[,c(1,2,5,10,11,13,12,16)] %>% #I assume Nfilt is the same as Nfail. What is it used for? Can I maybe just exclude it?
  mutate(across(!V1, as.numeric))
colnames(sorted_df) <- c("chrom","start_position","score", "coverage", "methylation_frequency","canon", "mod", "fail")
sorted_df$sum.modcalled <- sorted_df$canon + sorted_df$mod # add column with sum of the canonical calls and modified calls

#dataset (hp1) with selected cloumns
hp1 <- hp1[,c(1,2,5,10,11,13,12,16)] %>%
  mutate(across(!V1, as.numeric))
colnames(hp1) <- c("chrom","start_position","score", "coverage", "methylation_frequency","canon", "mod", "fail")
hp1$sum.modcalled <- hp1$canon + hp1$mod

#dataset (hp2) with selected cloumns
hp2 <- hp2[,c(1,2,5,10,11,13,12,16)] %>%
  mutate(across(!V1, as.numeric))
colnames(hp2) <- c("chrom","start_position","score", "coverage", "methylation_frequency","canon", "mod", "fail")
hp2$sum.modcalled <- hp2$canon + hp2$mod
```



```{r Import gene positions, include=FALSE}
gene_pos <- read.delim(paste(snakemake@input[["regions"]]), header=F)
colnames(gene_pos) <- c("chrom","start_position","end_position","gene_name","disorder","chrom_arm")
```



```{r Augment data, include=FALSE}
# Initialize empty data frame for appending
df = data.frame(matrix(nrow = 0, ncol = 11))

# Get a list of all the gene names
genes <- array(gene_pos$gene_name)

for (g in genes) {
  
  # Gene position
  d <- filter(gene_pos, gene_name == g)
  g_chrom <- pull(d, chrom)
  g_start <- pull(d, start_position)
  g_end <- pull(d, end_position)
  
  # Extract all sample data for this gene
  unphased_data <- filter(sorted_df, chrom == g_chrom & between(start_position, g_start, g_end))
  hp1_data <- filter(hp1, chrom == g_chrom & between(start_position, g_start, g_end))
  hp2_data <- filter(hp2, chrom == g_chrom & between(start_position, g_start, g_end))
  
  
  # Get positions with a control mean between 40% and 60%. 
  # These positions are more likely to be imprinted CpG positions
  c <- ctrls %>%
       filter(chrom == g_chrom & 
              between(start_position, g_start, g_end) & 
              between(methylation_frequency, 40, 60))
  
  # Extract those positions from sample data
  unphased_data_filtered <- semi_join(unphased_data, c, by = c("chrom", "start_position"))
  hp1_data_filtered <- semi_join(hp1_data, c, by = c("chrom", "start_position"))
  hp2_data_filtered <- semi_join(hp2_data, c, by = c("chrom", "start_position"))
  
  
  # Add to df to be printed in report
  df <- df %>%
         rbind(data.frame(d$disorder, 
                          d$chrom_arm, 
                          g, 
                          paste(d$start_position, d$end_position, sep="-"), 
                          nrow(unphased_data_filtered),
                          round(mean(unphased_data$sum.modcalled),0), 
                          round(mean(hp1_data$sum.modcalled),0), 
                          round(mean(hp2_data$sum.modcalled),0),
                          round(mean(unphased_data_filtered$methylation_frequency),1),
                          round(mean(hp1_data_filtered$methylation_frequency),1),
                          round(mean(hp2_data_filtered$methylation_frequency),1)))
}

colnames(df) <- c("Imprinting disorder",
                  "Chr", 
                  "Gene", 
                  "Genomic coordinates", 
                  "Number CpGs",
                  "Mean coverage",
                  "HP1 mean coverage",
                  "HP2 mean coverage", 
                  "Mean methylation (%)", 
                  "HP1 mean methylation (%)",
                  "HP2 mean methylation (%)")



df <- df %>%
  mutate(across(everything(), as.character)) %>%
  kable("html", escape = FALSE) %>%
  kable_styling() %>%
  # Highlight mean methylation below 25% and over 75% (and not NaN)
  column_spec(9, 
              color = ifelse(!is.na(df$"Mean methylation") & df$"Mean methylation" < 25 | !is.na(df$"Mean methylation") & df$"Mean methylation" > 75, "red", "black"),
              bold = ifelse(!is.na(df$"Mean methylation") & df$"Mean methylation" < 25 | !is.na(df$"Mean methylation") & df$"Mean methylation" > 75, TRUE, FALSE))
```



```{r print data frame, echo=FALSE}
df 
```



```{r Plot functions, echo=FALSE}
###################################
# Function for creating plot from unphased data
plot_unphased = function(g, g_chrom, g_start, g_end) {
  
  # Extract data for gene
  unphased_data <- filter(sorted_df, chrom == g_chrom & between(start_position, g_start, g_end))
  ctrl_data <- filter(ctrls, chrom == g_chrom & between(start_position, g_start, g_end))
  
  # If data frame is not empty; make plot
  if (nrow(unphased_data) > 0) {
    ggplot(data = unphased_data, aes(x = start_position, group = 1, se = F)) +
    geom_smooth(aes(y = methylation_frequency), colour = "black", se = F) +
    geom_point(aes(y = methylation_frequency), colour = "black", alpha = 0.3) +
    geom_ribbon(data = ctrl_data, aes(ymin = methylation_frequency - sd, ymax = methylation_frequency + sd), alpha = 0.2) +
    geom_ribbon(data = ctrl_data, aes(ymin = methylation_frequency - 1.5*sd, ymax = methylation_frequency + 0.1*sd), alpha = 0.1) + 
    labs(x = "Position on chromosome",
         y = "Methylation frequency (%)", 
         title = paste("Unphased: ", g, sep ="")) +
    coord_cartesian(xlim = c(g_start, g_end), ylim = c(-5,105), expand = F) + 
    theme_light() +
    theme(text = element_text(size = 15))
    # Empty data frame; make empty plot
  } else {
    ggplot() + 
      geom_blank() +
      labs(x = "Position on chromosome",
           y = "Methylation frequency (%)", 
           title = paste("Unphased: ", g, sep ="")) +
      coord_cartesian(xlim = c(g_start, g_end), ylim = c(-5,105), expand = F) + 
      theme_light() +
      theme(text = element_text(size = 15))
  }
}




###################################
# Function to create plot from phased data
plot_phased <- function(g, g_chrom, g_start, g_end) {
  
  # Extract data for gene
  hp1_data <- filter(hp1, chrom == g_chrom & between(start_position, g_start, g_end))
  hp2_data <- filter(hp2, chrom == g_chrom & between(start_position, g_start, g_end))
  
  # If data frame is not empty; make plot
  if (nrow(hp1_data) > 0 && nrow(hp2_data) > 0) {
    ggplot() +
        geom_smooth(data = hp1_data, aes(x = start_position, y = methylation_frequency), colour = "red", se = F) + 
        geom_point(data = hp1_data, aes(x = start_position, y = methylation_frequency), colour = "red", alpha = 0.3) +
        geom_smooth(data = hp2_data, aes(x = start_position, y = methylation_frequency), colour = "blue", se = F) + 
        geom_point(data = hp2_data, aes(x = start_position, y = methylation_frequency), colour = "blue", alpha = 0.3) +
        labs(x = "Position on chromosome",
             y = "Methylation frequency (%)", 
             title = paste("Phased: ", g, sep ="")) +
        coord_cartesian(xlim = c(g_start, g_end), ylim = c(-5,105), expand = F) + 
        theme_light() +
        theme(text = element_text(size = 15))  
  # Empty data frame; make empty plot
  } else {
    ggplot() + 
      geom_blank() +
      labs(x = "Position on chromosome",
           y = "Methylation frequency (%)", 
           title = paste("Phased: ", g, sep ="")) +
      coord_cartesian(xlim = c(g_start, g_end), ylim = c(-5,105), expand = F) + 
      theme_light() +
      theme(text = element_text(size = 15)) 
  }
}
```


## Visualization of methylation at imprinted regions
In the left side each imprinted region is shown with methylation frequency (%). Grey shaded regions mark mean $\pm$ 1 SD (darker grey) and $\pm$ 1.5 SD (lighter grey) for control samples. Haplotype phased methylation frequencies for the regions are seen to the right. HP1 is **<span style="color:red">red</span>** and HP2 is **<span style="color:blue">blue</span>**. If plots are empty phasing was not successful. 


```{r, fig.show="hold", out.width="50%", echo=FALSE, message=FALSE}
for (g in genes){
  d <- filter(gene_pos, gene_name == g)
  g_chrom <- pull(d, chrom)
  g_start <- pull(d, start_position)
  g_end <- pull(d, end_position)
  
  p1 <- plot_unphased(g, g_chrom, g_start, g_end)
  p2 <- plot_phased(g, g_chrom, g_start, g_end)
  
  print(p1)
  print(p2)
}
```